---
title: "smart_seq_analysis"
output: html_document
---

Data upload:
```{r}
barcodes <- read.csv("barcode_annotation.txt", sep = "", header = TRUE )
data <- readRDS("~/Documents/Sequencing/Smart_seq/Xpress.dgecounts.rds")
plate_map <- read.csv("~/Documents/Sequencing/Smart_seq/Plate_map.csv", sep = ";", header = TRUE)
```
Libraries
```{r}
library(dplyr)
library(ggplot2)
library(umap)
library(DESeq2)
#library(ensembldb)
#library(AnnotationHub)
library(biomaRt)
library(patchwork)
```


Filter reads for greater than 0.5 expression
```{r}
expression_data <- as.data.frame(data[["rpkm"]][["exon"]][["all"]]) %>%
  filter(rowSums(.) >= 2500)
```
Rename columns using a mapping
```{r}
expression_data <- expression_data %>%
  rename_with(~ barcodes$WellID[match(., barcodes$XC_DNBPE_TS)], .cols = everything())
```

PCA plot
```{r}
# Transpose so that samples are rows and genes are columnsß
expression_transposed <- t(expression_data)
pca_result <- prcomp(expression_transposed, scale. = TRUE)
pca_df <- data.frame(Well = rownames(pca_result$x),  # Extract sample names
                     PC1 = pca_result$x[,1], 
                     PC2 = pca_result$x[,2])

# Merge PCA data with category info
pca_df <- merge(pca_df, plate_map, by = "Well")

# Create PCA plot
sample_type_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = Sample_type)) +
  geom_point(size = 3) +
  labs(title = "PCA Plot", x = "PC1", y = "PC2") +
  theme_minimal()
sample_type_PCA
```
Making a uMAP
```{r}
# Run UMAP
umap_result <- umap(expression_transposed)

# Convert UMAP results into a data frame
umap_df <- data.frame(Well = rownames(expression_transposed),
                      UMAP1 = umap_result$layout[,1], 
                      UMAP2 = umap_result$layout[,2])
# Merge UMAP data with category info
umap_df <- merge(umap_df, plate_map, by = "Well")

#make the plot
sample_type_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = Sample_type)) +
  geom_point(size = 3) +
  labs(title = "UMAP Plot of Gene Expression", x = "UMAP1", y = "UMAP2") +
  theme_minimal()
sample_type_umap
```
```{r}
umap_df_filtered <- subset(umap_df, Sample_type %in% c("Blood_CSP", "BM_CSP"))

sample_type_umap <- ggplot(umap_df_filtered, aes(x = UMAP1, y = UMAP2, color = Sample_type)) +
  geom_point(size = 3) +
  labs(x = "UMAP1", y = "UMAP2") +
  theme_bw() +
  scale_color_lancet()  +
  theme(aspect.ratio = 1) # Optional if you want consistent colors

sample_type_umap

#save
ggsave("~/Documents/Sequencing/Figures/umap_plot.png", plot = sample_type_umap, width = 8, height = 6)
```

#checking colSums of control columns
```{r}
col_sums_df <- data.frame(t(colSums(expression_data[, c("H23", "H24", "J1","J2","J3","J4","J5","J6","J7","J8","J9","J10")])))
rowMeans(col_sums_df)
```

```{r}
# View top genes for PC1
# Extract loadings (gene contributions)
loadings <- pca_result$rotation  # Each row is a gene, each column is a PC

# View top genes for PC1
top_genes_PC1 <- head(loadings[order(abs(loadings[,1]), decreasing = TRUE), ], 1000)
print(top_genes_PC1)

top_genes_list <- lapply(1:3, function(i) {
  top_genes <- head(loadings[order(abs(loadings[,i]), decreasing = TRUE), ], 1000)
  return(data.frame(Gene = rownames(top_genes), Contribution = top_genes[,i], PC = paste0("PC", i)))
})

top_genes_df <- do.call(rbind, top_genes_list)  # Combine results into a data frame
print(top_genes_df)

ggplot(top_genes_df, aes(x = reorder(Gene, abs(Contribution)), y = Contribution, fill = PC)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  theme_minimal() +
  labs(title = "Top Genes Contributing to PCA Variation", x = "Genes", y = "Contribution")

```
```{r}
# View top genes for PC1
# Extract loadings (gene contributions)
loadings <- pca_result$rotation  # Each row is a gene, each column is a PC

# View top genes for PC1
top_genes_PC2 <- head(loadings[order(abs(loadings[,2]), decreasing = TRUE), ], 1000)
print(top_genes_PC2)

top_genes_list_PC2 <- lapply(1:3, function(i) {
  top_genes_PC2 <- head(loadings[order(abs(loadings[,i]), decreasing = TRUE), ], 1000)
  return(data.frame(Gene = rownames(top_genes_PC2), Contribution = top_genes_PC2[,i], PC = paste0("PC", i)))
})

top_genes_df_PC2 <- do.call(rbind, top_genes_list_PC2)  # Combine results into a data frame
print(top_genes_df)

ggplot(top_genes_df_PC2, aes(x = reorder(Gene, abs(Contribution)), y = Contribution, fill = PC)) +
  geom_bar(stat = "identity") +
  coord_flip() + 
  theme_minimal() +
  labs(title = "Top Genes Contributing to PCA Variation", x = "Genes", y = "Contribution")

```
Annotating genes from the ensembl database
```{r}
library(biomaRt)
ensembl <- useEnsembl(biomart = "genes", dataset = "mfascicularis_gene_ensembl")
gene_info <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "description"),
                   filters = "ensembl_gene_id",
                   values = c("ENSMFAG00000013144",	"ENSMFAG00000038421", "ENSMFAG00000008667"),
                   mart = ensembl)
print(gene_info)
```
```{r}
ensembl <- useEnsembl(biomart = "genes", dataset = "mfascicularis_gene_ensembl")
# View the available GO terms for B cell-related processes
go_terms <- c("GO:0006955")  # Example GO terms for immune response
# Fetch genes related to the specified GO terms
bcell_genes <- getBM(attributes = c("ensembl_gene_id", "external_gene_name", "description"),
                     filters = "go_parent_term",  # Use GO term to filter
                     values = go_terms,  # Provide the GO term(s)
                     mart = ensembl)
listFilters(ensembl)
# View the results
print(bcell_genes)
```
```{r}
matched_genes <- merge(top_genes_list[[1]], bcell_genes, by.x = "Gene", by.y = "ensembl_gene_id")

```
Try PCA for WASP
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
wasp_expression <- t(expression_data["ENSMFAG00000034169", ])  # if 'WASP' is the gene name
pca_result <- prcomp(t(expression_data), scale. = TRUE)
# Create a data frame with the PCA results
pca_df <- as.data.frame(pca_result$x)

# Add the expression values of WASP as a column to the PCA data frame
pca_df$WASP_expression <- wasp_expression

# Load ggplot2 package
library(ggplot2)

# Create the PCA plot with coloring by WASP expression
ggplot(pca_df, aes(x = PC1, y = PC2, color = WASP_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by WASP Expression", x = "PC1", y = "PC2") +
  theme_minimal()

```
uMAP
```{r}
# Transpose the expression data (samples as rows, genes as columns)
expression_transposed <- t(expression_data)

# Perform UMAP
umap_result <- umap(expression_transposed)
# Create a data frame for plotting
umap_df <- as.data.frame(umap_result$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

wasp_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = wasp_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP Colored by WASP Expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

wasp_expresion_umap
```

B cell specific genes
-CD19
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CD19_expression <- t(expression_data["ENSMFAG00000003219", ])  # if 'WASP' is the gene name
pca_result <- prcomp(t(expression_data), scale. = TRUE)
# Create a data frame with the PCA results
pca_df <- as.data.frame(pca_result$x)

# Add the expression values of WASP as a column to the PCA data frame
pca_df$CD19_expression <- CD19_expression

# Create the PCA plot with coloring by WASP expression
CD19_PCA <-ggplot(pca_df, aes(x = PC1, y = PC2, color = CD19_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by CD19 Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
# Perform UMAP
umap_result <- umap(expression_transposed)
# Create a data frame for plotting
umap_df <- as.data.frame(umap_result$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

CD19_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = CD19_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP Colored by CD19 Expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

CD19_expresion_umap
```
-IGHM
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
IGHM_expression <- t(expression_data["ENSMFAG00000054250", ])  # if 'WASP' is the gene name
pca_result <- prcomp(t(expression_data), scale. = TRUE)
# Create a data frame with the PCA results
pca_df <- as.data.frame(pca_result$x)

# Add the expression values of WASP as a column to the PCA data frame
pca_df$IGHM_expression <- IGHM_expression

# Create the PCA plot with coloring by WASP expression
IGHM_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = IGHM_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by IGHM Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
# Perform UMAP
umap_result <- umap(expression_transposed)
# Create a data frame for plotting
umap_df <- as.data.frame(umap_result$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

IGHM_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = IGHM_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP Colored by IGHM Expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

IGHM_expresion_umap
```
CD3D
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CD3D_expression <- t(expression_data["ENSMFAG00000045644", ])  # if 'WASP' is the gene name
pca_result <- prcomp(t(expression_data), scale. = TRUE)
# Create a data frame with the PCA results
pca_df <- as.data.frame(pca_result$x)

# Add the expression values of WASP as a column to the PCA data frame
pca_df$CD3D_expression <- CD3D_expression

# Create the PCA plot with coloring by WASP expression
CD3D_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = CD3D_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by CD3D Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```

CD3E
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CD3E_expression <- t(expression_data["ENSMFAG00000040556", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$CD3E_expression <- CD3E_expression

# Create the PCA plot with coloring by WASP expression
CD3E_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = CD3E_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by CD3E Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
CD3G
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CD3G_expression <- t(expression_data["ENSMFAG00000040063", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$CD3G_expression <- CD3E_expression

# Create the PCA plot with coloring by WASP expression
CD3G_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = CD3G_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by CD3G Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
CD40
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CD40_expression <- t(expression_data["ENSMFAG00000032922", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$CD40_expression <- CD40_expression

# Create the PCA plot with coloring by WASP expression
CD40_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = CD40_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by CD40 Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
# Perform UMAP
umap_result <- umap(expression_transposed)
# Create a data frame for plotting
umap_df <- as.data.frame(umap_result$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

CD40_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = CD40_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP Colored by CD40 Expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

CD40_expresion_umap
```
CXCR3
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CXCR3_expression <- t(expression_data["ENSMFAG00000031295", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$CXCR3_expression <- CXCR3_expression

# Create the PCA plot with coloring by WASP expression
CXCR3_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = CXCR3_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by CD3G Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
CXCR3_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = CXCR3_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP colored by CXCR3 Expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

CXCR3_expresion_umap
```
CXCR4
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CXCR4_expression <- t(expression_data["ENSMFAG00000037063", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$CXCR4_expression <- CXCR4_expression

# Create the PCA plot with coloring by WASP expression
CXCR4_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = CXCR4_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by CXCR4 Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
CXCR4_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = CXCR4_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP colored by CXCR3 Expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

CXCR4_expresion_umap
```
CD80
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CD80_expression <- t(expression_data["ENSMFAG00000032987", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$CD80_expression <- CD80_expression

# Create the PCA plot with coloring by WASP expression
CD80_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = CD80_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by CD80 Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
CD80_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = CD80_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP colored by CD80 expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

CD80_expresion_umap
```
CCR6
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
CCR6_expression <- t(expression_data["ENSMFAG00000062524", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$CCR6_expression <- CCR6_expression

# Create the PCA plot with coloring by WASP expression
CCR6_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = CCR6_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot colored by CCR6 expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
CCR6_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = CCR6_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP colored by CCR6 expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

CCR6_expresion_umap
```
JCHAIN
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
JCHAIN_expression <- t(expression_data["ENSMFAG00000065046", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$JCHAIN_expression <- JCHAIN_expression

# Create the PCA plot with coloring by WASP expression
JCHAIN_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = JCHAIN_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA plot colored by JCHAIN expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
JCHAIN_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = JCHAIN_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP colored by CCR6 expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

JCHAIN_expresion_umap
```
AICDA
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
AICDA_expression <- t(expression_data["ENSMFAG00000064606", ])  
# Add the expression values of WASP as a column to the PCA data frame
pca_df$AICDA_expression <- AICDA_expression

# Create the PCA plot with coloring by WASP expression
AICDA_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = AICDA_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA plot colored by AICDA expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
AICDA_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = AICDA_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP colored by AICDA expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

AICDA_expresion_umap
```
All plots on the same figure
```{r}
#PCA plots
all_PCA_plots <- (CD19_PCA | IGHM_PCA | CD3D_PCA) / (CD3E_PCA | CD3G_PCA | CD40_PCA) / (CXCR3_PCA | CXCR4_PCA | CD80_PCA) / (CCR6_PCA | JCHAIN_PCA |AICDA_PCA) +
  plot_layout(guides = "collect") & theme(legend.position = "right")
all_PCA_plots
```
```{r}
#PCA plots
PCA_plots <- (CD19_PCA | IGHM_PCA | CD40_PCA) / (CXCR3_PCA | CXCR4_PCA | CD80_PCA) / (CCR6_PCA | JCHAIN_PCA |AICDA_PCA) +
  plot_layout(guides = "collect") & theme(legend.position = "none")
PCA_plots
```
uMAP
```{r}
#PCA plots
umap_plots <- (CD19_expresion_umap | IGHM_expresion_umap | CD40_expresion_umap) / (CXCR3_expresion_umap | CXCR4_expresion_umap | CD80_expresion_umap) / (CCR6_expresion_umap | JCHAIN_expresion_umap |AICDA_expresion_umap) +
  plot_layout(guides = "collect") & theme(legend.position = "none")
umap_plots
```

PCA and umap for only blood CSP and BM CSP
```{r}
# Filter the sample IDs in df_sample_types for 'blood' or 'bone marrow'
filtered_plate_map <- plate_map %>%
  filter(Sample_type %in% c("Blood_CSP", "BM_CSP")) %>%
  pull(Well)

# Subset the df_gene_counts dataframe based on the filtered sample IDs
CSP_expression_transposed<- as.data.frame(expression_transposed) %>%
  filter(rownames(expression_transposed) %in% filtered_plate_map)

  
#filter if rownames(expression_transposed) are of Sample_type Blood_CSP and BM_CSP
# Calculate the variance of each column
column_variances <- apply(CSP_expression_transposed, 2, var)

# Identify columns with zero variance
constant_columns <- which(column_variances == 0)

# Remove the constant columns
CSP_expression_transposed_filtered <- CSP_expression_transposed[, -constant_columns]

# Now run PCA on the filtered data
CSP_pca_result <- prcomp(CSP_expression_transposed_filtered, scale. = TRUE)

CSP_pca_result <- prcomp(CSP_expression_transposed_filtered, scale. = TRUE)
CSP_pca_df <- data.frame(Well = rownames(CSP_pca_result$x),  # Extract sample names
                     PC1 = CSP_pca_result$x[,1], 
                     PC2 = CSP_pca_result$x[,2])

# Merge PCA data with category info
CSP_pca_df <- merge(CSP_pca_df, plate_map, by = "Well")

# Create PCA plot
CSP_sample_type_PCA <- ggplot(CSP_pca_df, aes(x = PC1, y = PC2, color = Sample_type)) +
  geom_point(size = 3) +
  labs(title = "PCA Plot", x = "PC1", y = "PC2") +
  theme_minimal()
CSP_sample_type_PCA
```
uMAP for CSP-specific only
```{r}
# Run UMAP
CSP_umap_result <- umap(CSP_expression_transposed)

# Convert UMAP results into a data frame
CSP_umap_df <- data.frame(Well = rownames(CSP_expression_transposed),
                      UMAP1 = CSP_umap_result$layout[,1], 
                      UMAP2 = CSP_umap_result$layout[,2])
# Merge UMAP data with category info
CSP_umap_df <- merge(CSP_umap_df, plate_map, by = "Well")

#make the plot
CSP_sample_type_umap <- ggplot(CSP_umap_df, aes(x = UMAP1, y = UMAP2, color = Sample_type)) +
  geom_point(size = 3) +
  labs(title = "UMAP Plot of Gene Expression", x = "UMAP1", y = "UMAP2") +
  theme_minimal()
CSP_sample_type_umap
```
B cell specific genes
-CD19
```{r}
CSP_expression_data <- t(CSP_expression_transposed_filtered)
CSP_CD19_expression <- CSP_expression_data["ENSMFAG00000003219", ]  # if 'WASP' is the gene name

# Add the expression values of WASP as a column to the PCA data frame
CSP_pca_df$CD19_expression <- CSP_CD19_expression

# Create the PCA plot with coloring by WASP expression
CSP_CD19_PCA <-ggplot(CSP_pca_df, aes(x = PC1, y = PC2, color = CSP_CD19_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "CD19 Expression for CSP only", x = "PC1", y = "PC2") +
  theme_minimal()
CSP_CD19_PCA
```
uMAP
```{r}

CSP_CD19_expresion_umap <- ggplot(CSP_umap_df, aes(x = UMAP1, y = UMAP2, color = CSP_CD19_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP Colored by CD19 Expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

CSP_CD19_expresion_umap
```
-IGHM (do this for the rest of the graphs!!!)
```{r}
# Extract the expression of the WASP gene (replace 'WASP' with the actual gene name or ID in your dataset)
IGHM_expression <- t(expression_data["ENSMFAG00000054250", ])  # if 'WASP' is the gene name
pca_result <- prcomp(t(expression_data), scale. = TRUE)
# Create a data frame with the PCA results
pca_df <- as.data.frame(pca_result$x)

# Add the expression values of WASP as a column to the PCA data frame
pca_df$IGHM_expression <- IGHM_expression

# Create the PCA plot with coloring by WASP expression
IGHM_PCA <- ggplot(pca_df, aes(x = PC1, y = PC2, color = IGHM_expression)) +
  geom_point(size = 1) +  # You can adjust the size of the points
  scale_color_gradient(low = "blue", high = "red") +  # Set color scale (you can adjust colors)
  labs(title = "PCA Plot Colored by IGHM Expression", x = "PC1", y = "PC2") +
  theme_minimal()
```
uMAP
```{r}
# Perform UMAP
umap_result <- umap(expression_transposed)
# Create a data frame for plotting
umap_df <- as.data.frame(umap_result$layout)
colnames(umap_df) <- c("UMAP1", "UMAP2")

IGHM_expresion_umap <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = IGHM_expression)) +
  geom_point(size = 3) +  # Scatter plot with points sized at 3
  scale_color_gradient(low = "blue", high = "red") +  # Color scale (adjust as needed)
  labs(title = "UMAP Colored by IGHM Expression", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

IGHM_expresion_umap
```
Try to map in regards to IGH expression
```{r}
library(dplyr)
library(tibble)

IGH_data_smartseq <- read.csv("~/Documents/Sequencing/Smart_seq/BCR_analysis/smartseq_airr.tsv", sep = "\t")
#Filter for onlyIGH
IGH_data_smartseq <- IGH_data_smartseq %>%
  filter(grepl("IGH", locus)) %>%
  mutate(WellID = sub("_.*", "", sequence_id)) %>%
  mutate(v_family = sub("-.*", "", v_call)) %>%
  mutate(d_family = sub("-.*", "", d_call)) %>%
  mutate(j_family = sub("-.*", "", j_call))

expression_transposed_for_IGH <- as.data.frame(expression_transposed) %>%
    rownames_to_column(var = "WellID")%>%
    left_join( IGH_data_smartseq, by = "WellID") 

uMAP_expression_transposed_for_IGH <- expression_transposed_for_IGH %>%
    dplyr::select(where(is.numeric))

# Run UMAP
umap_result <- umap(uMAP_expression_transposed_for_IGH )

# Combine UMAP results with metadata
umap_df <- as.data.frame(umap_result$layout) %>%
  mutate(
    WellID = expression_transposed_for_IGH$WellID,
    v_family = expression_transposed_for_IGH$v_family
  )
colnames(umap_df)[1:2] <- c("UMAP1", "UMAP2")

```

